%
% $Header$
%
\section{Input/Output Semantics}
\label{io-semantics}
\index{input/output!semantics}

The behaviour of a \Haskell{} program performing I/O is given within
the environment in which it is running.  That environment will be described
using standard \Haskell{} code augmented with a non-deterministic
merge operator.

The state of the operating system (OS state) that is relevant to
\Haskell{} programs is completely described by the file system and the
channel system.  The channel system is split into two subsystems, the
input channel system and the output channel system.
\bprog
\mbox{\tt type\ State\ =\ (FileSystem,\ ChannelSystem)}\\
\mbox{\tt type\ FileSystem\ \ \ \ =\ String\ ->\ Response}\\
\mbox{\tt type\ ChannelSystem\ =\ (ICs,\ OCs)}\\
\mbox{\tt type\ ICs\ \ \ =\ String\ ->\ (Agent,\ Open)}\\
\mbox{\tt type\ OCs\ \ \ =\ String\ ->\ Response}\\
\mbox{\tt type\ Agent\ =\ (FileSystem,\ OCs)\ ->\ Response}\\
\mbox{\tt type\ Open\ \ =\ PId\ ->\ Bool}\\
\mbox{\tt type\ PId\ \ \ =\ Int}\\
\mbox{\tt type\ PList\ =\ [(PId,[Request->Response])]}
\eprog
An agent maps a list of OS states to responses.  Those
responses will be used as the contents of input channels, and
thus can depend on output channels, other input channels, files, or
any combination thereof.  For example, a valid implementation must
allow the user to act as agent between the standard output
channel and standard input channel.

Each running process (i.e.~program) has a unique \mbox{\tt PId}.
Elements of \mbox{\tt PList} are lists of running programs.

\bprog
\mbox{\tt os\ ::\ TagReqList\ ->\ State\ ->\ (TagRespList,\ State)}\\
\mbox{\tt type\ TagRespList\ =\ [(PId,Response)]}\\
\mbox{\tt type\ TagReqList\ \ =\ [(PId,Request)]}
\eprog
The operating system is modeled as a (non-deterministic) function
\mbox{\tt os}. The \mbox{\tt os} takes a tagged request list and an initial state, and
returns a tagged response list and a final state.  Given a list of
programs \mbox{\tt pList}, \mbox{\tt os} must exhibit this behaviour:
\bprog
\mbox{\tt (tagRespList,\ state')\ =\ os\ tagReqList\ state}\\
\mbox{\tt tagReqList\ =\ merge\ [\ zip\ [pId,pId..]\ (proc\ (untag\ pId\ tagRespList))}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ |\ (pId,\ proc)\ <-\ pList\ ]}
\eprog
where \mbox{\tt merge} is a non-deterministic merge of a list of lists, and
\mbox{\tt untag} is:
\bprog
\mbox{\tt untag\ n\ []\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ =\ []}\\
\mbox{\tt untag\ n\ ((m,resp):resps)\ =\ if\ n==m\ then\ resp:(untag\ n\ resps)}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ else\ untag\ n\ resps}
\eprog
This relationship can be generalised to
include requests such as \mbox{\tt CreateProcess}.

A valid implementation must ensure that the input channel
system is defined at \mbox{\tt stdin} and the output channel system is
defined at \mbox{\tt stdout}, \mbox{\tt stderr}, and \mbox{\tt stdecho}.  If the agent
attached to standard input is called \mbox{\tt user} (i.e.~\mbox{\tt ics\ stdin} has
form \mbox{\tt (user,\ open)}), then \mbox{\tt user} must depend at least on
standard output.  In other words, this constraint must hold:
\bprog
\mbox{\tt user\ [...,\ (fs,(ics,ocs)),\ ...]\ =\ ...\ user'\ (ocs\ stdout)\ ...}
\eprog
where \mbox{\tt user'} is a {\em strict}, but otherwise arbitrary, function
modelling the user.  Its strictness corresponds to the user's
consumption of standard output whilst determining
standard input.

The rest of this section specifies the required behaviour of \mbox{\tt os} in
response to each kind of request.  This semantics is relatively
abstract and omits any reference to hardware errors (e.g.~``bad
sector on disk'') and system dependent errors (e.g.~``access rights
violation'').  Implementation-specific requests (for example the
environment requests) are not shown here.  We describe only
the text version of the requests: the binary version differs
trivially. \mbox{\tt os} is defined by:
\bprog
\mbox{\tt os\ ::\ TagReqList\ ->\ State\ ->\ (TagRespList,State)}\\
\mbox{\tt }\\[-8pt]
\mbox{\tt os\ []\ state\ =\ ([],\ state)}\\
\mbox{\tt os\ ((n,\ ReadChan\ name):es)\ state@(fs,(ics,ocs))\ =}\\
\mbox{\tt \ \ \ \ (alist',state')\ where}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ (agent,open)\ =\ ics\ name}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ alist'\ =\ (n,\ (if\ open\ n}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ then\ fail}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ else\ (agent\ (fs,ocs))\ ))\ :\ alist}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ fail\ =\ Failure\ (OtherError\ "Channel\ already\ open{\char'134}n")}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ (alist,state')\ =\ os\ es\ (fs,\ (update\ ics\ name}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ (agent,\ update\ open\ n\ true),}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ ocs))}
\eprog
where the auxiliary function \mbox{\tt update} is defined by:
\bprog
\mbox{\tt update\ f\ x\ v\ x'\ =\ if\ x==x'\ then\ v\ else\ f\ x}
\eprogNoSkip

If an attempt is made to read a non-existent channel, \mbox{\tt ics}
returns an agent that gives the appropriate error message when
applied to its arguments.  This definition is generalised in the
obvious way for the behaviour of \mbox{\tt ReadChannels}.  In particular, \mbox{\tt ack}
must be created by non-deterministically merging the result of applying
each agent to the stream of future states.
%Applying
%agents to the entire system state is simply meant to capture the most
%general situation where an adept agent might be able to
%observe the entire state of the system.
\bprog
\mbox{\tt os\ ((n,\ AppendChan\ name\ contents):es)\ state@(fs,(ics,ocs))\ =}\\
\mbox{\tt \ \ \ \ (alist',state')\ where}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ alist'\ =\ ack:alist}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ ack\ =}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ (n,}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ case\ (ocs\ name)\ of}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ Failure\ msg\ ->\ Failure\ (SearchError\ "Nonexistent\ Channel")}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ Str\ ochan\ ->\ Success}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ Bn\ ochan\ ->\ Failure\ (FormatError\ "format\ error")}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ )}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ (alist,state')\ =\ os\ es\ (fs,(ics,}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ case\ (ocs\ name)\ of}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Failure\ msg\ ->\ ocs}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Str\ ochan\ ->\ update\ ocs\ name}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ (Str\ (ochan\ ++\ contents))}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Bn\ ochan\ ->\ ocs}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ ))}
\eprogNoSkip
\bprog
\mbox{\tt os\ ((n,\ ReadFile\ name):es)\ state@(fs,(ics,ocs))\ =}\\
\mbox{\tt \ \ \ \ (alist',state')\ where}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ alist'\ =\ ack\ :\ alist}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ ack\ =\ (n,}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ case\ (fs\ name)\ of}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Failure\ msg\ ->\ Failure\ (SearchError\ "File\ not\ found")}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Str\ string\ ->\ Str\ string}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Bn\ binary\ ->\ Failure\ (FormatError\ "")}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ )}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ (alist,state')\ =\ os\ es\ state}\\
\mbox{\tt }\\[-8pt]
\mbox{\tt os\ ((n,\ WriteFile\ name\ contents):es)\ state@(fs,(ics,ocs))\ =}\\
\mbox{\tt \ \ \ \ (alist',state')\ where}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ alist'\ =\ (n,\ Success):alist}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ (alist,state')\ =\ os\ es\ (update\ fs\ name\ (Str\ contents),}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ (ics,ocs))}
\eprogNoSkip
\bprog
\mbox{\tt os\ ((n,\ AppendFile\ name\ contents):es)\ state@(fs,(ics,ocs))\ =}\\
\mbox{\tt \ \ \ \ (alist',state')\ where}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ alist'\ =\ ack:alist}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ ack\ =\ (n,\ }\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ case\ (fs\ name)\ of}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Failure\ msg\ ->\ Failure\ (SearchError\ "file\ not\ found")}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Str\ s\ ->\ Success}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Bn\ \ b\ ->\ Failure\ (FormatError\ "")}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ )}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ (alist,state')\ =\ os\ es\ (newfs,\ (ics,ocs))\ \ where}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ newfs\ =\ case\ (fs\ name)\ of}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Failure\ msg\ ->\ fs}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Str\ s\ ->}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ update\ fs\ name\ (Str\ (s++contents))}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Bn\ \ b\ ->\ fs}
\eprogNoSkip
\bprog
\mbox{\tt os\ ((n,\ DeleteFile\ name):es)\ state@(fs,(ics,ocs))\ =}\\
\mbox{\tt \ \ \ \ (alist',state')\ where}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ alist'\ =\ ack\ :\ alist}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ ack\ =\ (n,}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ case\ (fs\ name)\ of}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Failure\ msg\ ->\ Failure\ (SearchError\ "file\ not\ found")}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Str\ s\ ->\ Success}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Bn\ b\ ->\ Success}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ )}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ (alist,state')\ =\ os\ es\ (case\ (fs\ name)\ of}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Failure\ msg\ ->\ fs}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Str\ s\ ->\ update\ fs\ name\ fail}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Bn\ b\ ->\ update\ fs\ name\ fail,}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ (ics,ocs))}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ fail\ =\ Failure\ (SearchError\ "file\ not\ found")}\\
\mbox{\tt }\\[-8pt]
\mbox{\tt os\ ((n,StatusFile\ name):es)\ state@(fs,(ics,ocs))\ =\ (alist',state')\ where}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ alist'\ =\ ack\ :\ alist}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ ack\ =\ (n,}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ case\ (fs\ name)\ of}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Failure\ msg\ ->\ Failure\ (SearchError\ "File\ not\ found")}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Str\ string\ ->\ Str\ "t"++(rw\ n\ fs\ name)}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Bn\ binary\ ->\ Str\ "b"++(rw\ n\ fs\ name)}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ )}\\
\mbox{\tt \ \ \ \ \ \ \ \ \ \ (alist,\ state')\ =\ os\ es\ state}
\eprog
where \mbox{\tt rw} is a function that determines the read and write status
of a file for this particular process.

% Local Variables: 
% mode: latex
% End:
